Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-02-05 WSSSPE1_report/WSSSPE1_summary_report.tex
Update WSSSPE1_summary_report.tex

\usepackage{color}
\usepackage{dcolumn}
\usepackage{float}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{subfigure}
\usepackage{psfrag}
\usepackage{tabularx}
\usepackage[hyphens]{url}
\usepackage{wrapfig}
\usepackage{enumitem}

@@

\usepackage{array}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

\note{papers and presentations and URLs should be references - see
  figshare-web in .bib file for URL style example}

@@

The First Workshop on Sustainable Software for Science: Practice and
Experiences (WSSSPE1,
\url{http://wssspe.researchcomputing.org.uk/WSSSPE1}) was held on
Sunday, 17 November 2013, in conjunction with the 2013 International
Conference for High Performance Computing, Networking, Storage and
Analysis (SC13, \url{http://sc13.supercomputing.org}).

Because progress in scientific research is dependent on the quality
and accessibility of software at all levels, it is now critical to
address many challenges related to the development, deployment, and
maintenance of reusable software. In addition, it is essential that
scientists, researchers, and students are able to learn and adopt
software-related skills and methodologies. Established researchers are
already acquiring some of these skills, and in particular a
specialized class of software developers is emerging in academic
environments who are an integral and embedded part of successful
research teams. The WSSSPE workshop was intended to provide a forum
for discussion of the challenges, including both positions and
experiences.  The short papers and discussion have been archived to
provide a basis for continued discussion, and the workshop has fed
into the collaborative writing of this journal publication.
Additional papers based on extended versions of workshop submissions
are expected.  The level of interest in the workshop has led the
organizers, working with some of the submitters and attendees, to plan
two additional workshops (at the 2014 SciPy and SC conferences) and to
turn the workshop website into a community website that can be used as
a focus for further discussion and progress.  A minisymposium on
``Reliable Computational Science'' that aims to further explore some
of the key issues raised in WSSSPE is co-organized by a participant in
the 2014 SIAM Annual Meeting.

Before the WSSSPE1 workshop took place, the organizers self-published
a paper~\cite{WSSSPE1-pre-report} to document the process of
organizing and advertising the workshop, collecting and reviewing the
papers, and putting together the agenda.  Section~\ref{sec:process} is
based on that paper.  The remainder of this paper is based on the
workshop itself, as documented in a set of collaborative notes taken
during the workshop~\cite{WSSSPE1-google-notes}, including discussion
about the keynote presentations (\S\ref{sec:keynotes}), the three
panels (\S\ref{sec:devel}-\ref{sec:community}), other discussion
(\S\ref{sec:other}), cross-cutting issues (\S\ref{sec:cross-cutting}),
use cases (\S\ref{sec:use-cases}), and conclusions
(\S\ref{sec:conclusions}).

\section{Workshop Process and Agenda} \label{sec:process}

WSSSPE1 was organized in collaboration of the relatively small group
of five organizers and a larger peer-review committee with 36
members. This committee had early influence on the organization, e.g.,
on the specific call for papers.

@@

\begin{itemize}[leftmargin=0.2in]
\item the development process that leads to new (versions of) software
\begin{itemize}[leftmargin=0.2in]
\item how fundamental research in computer science or
  science/engineering domains is turned into reusable software
\item software created as a by-product of research
\item impact of computer science research on the development of
  scientific software
\end{itemize}
\item the support and maintenance of existing software, including
  software engineering
\begin{itemize}[leftmargin=0.2in]
\item governance, business, and sustainability models
\item the role of community software repositories, their operation and
  sustainability
\end{itemize}
\item the role of open source communities or industry
\item use of the software
\begin{itemize}[leftmargin=0.2in]
\item growing communities
\item reproducibility, transparency needs that may be unique to science
\end{itemize}
\item policy issues, such as
\begin{itemize}[leftmargin=0.2in]
\item measuring usage and impact
\item software credit, attribution, incentive, and reward
\item career paths for developers and institutional roles
\item issues related to multiple organizations and countries,
  such as intellectual property, licensing, etc.
\item mechanisms and venues for publishing software, and the role of
  publishers
\end{itemize}
\item education and training
\end{itemize}

@@

58 submissions were received, and almost all submitters used either
arXiv~\cite{arXiv-web} or figshare~\cite{figshare-web} to self-publish
their papers.  A peer review process followed the submissions, where
the 58 papers received 181 reviews, an average of 3.12 reviews per
paper. Reviews were completed using a Google form, which allowed
reviewers to choose papers they wanted to review, then to provide
scores on their relevance and general comments to the organizers and
to the authors. The review reports enabled the organizers to decide
which papers to associate with the workshop, and allowed the authors
to improve their papers.

The organizers decided to list 54 of the papers as significant
contributions to the workshop, a very high acceptance rate, but one
that is reasonable, given the goal of broad participation and the fact
that the reports were already self-published. The papers were
grouped into three main categories, namely \emph{Developing and
  Supporting Software}, \emph{Policy}, and \emph{Communities}. Each
subject area was associated with a panel and discussion at the
workshop.

The workshop itself consisted of two keynote presentations and the
three panels/discussion sessions. The panels were organized based on a
classification of the workshop submissions into three categories,
following the themes of the call for papers as modified by the areas
of the submissions. Three to four representatives from each submission
category were appointed as panelists, and assigned to read a subset of
the paper in that category and then discuss them in the panel.

\section{Keynotes} \label{sec:keynotes}

The WSSSPE1 workshop began with two keynote presentations.

@@

He noted that in places, development in science has improved thanks to
open source and Github, but for the most part remains arcane. He
argues that there is a lot we can learn from the app-store model. He
mentioned BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}}
and
OpenScienceDataCloud\footnote{\url{https://www.opensciencedatacloud.org/}}
as distribution channels.

@@

He experienced problems with staff retention with respect to private
organizations which reward those combining research and software
expertise (the ``Google Bus''). However, he is a strong fan of private
sector sustainability sources and public-private partnerships. He
noted that making a successful business from scientific software alone
is very rare: founders overvalue while customers undervalue. He noted
that to last, an open source project needs a minimal funding
requirement even with a vibrant community -- good will only goes so
far if you are being paid to do something else.

He talked about grant schemes of relevance in the U.S., particularly
with regard to technology
transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}}

@@

\subsection{Scientific Software and the Open Collaborative Web, Arfon Smith}

@@

His talk used an example from data reduction in Astronomy, where he
needed to remove interfering effects from the device. He built a ``bad
pixel mask'', and realized that while this is persistent, there was no
way/practice of sharing this with the data, so many researchers
repeated the same calculations. He estimated that 13 person-years were
wasted in this redundant calculation.

@@

\section{Developing and Supporting Software} \label{sec:devel}

@@

What is sustainable? What are the best practices, can workshops fund
meetings to help define and improve best practices in these areas?
Software plans from the funding agencies? This topic is expanded upon
in \S\ref{sec:cross-cutting}, since it was discussed up in multiple
parts of the WSSSPE1 workshop.

@@

\section{Policy} \label{sec:policy}

@@

\section{Communities} \label{sec:community}

@@

@@

\section{Other Discussion} \label{sec:other}

@@

\section{Cross-cutting Issue: Defining Sustainability}  \label{sec:cross-cutting}

\item From discussion in Phil Bourne's keynote:
\begin{itemize}
\item Can sustainability be defined as the effort that happens to make
  the essential things continue?
\item What is sustainability? What is it that we want to sustain? Is
  what we want to sustain valuable? Who cares if it goes away and how
  do you measure how much they care?
\end{itemize}

\item from discussion in developing panel

\begin{itemize}
\item what does sustainable mean?
\begin{itemize}
\item from http://oss-watch.ac.uk/resources/ssmm: When choosing
  software for procurement or development reuse - regardless of the
  license and development model you will use - you need to consider
  the future. While a software product may satisfy today's needs, will
  it satisfy tomorrow's needs? Will the supplier still be around in
  five years' time? Will the supplier still care for all its customers
  in five years' time? Will the supplier be responsive to bug reports
  and feature requests? In other words, is the software sustainable?
\item I think this is a key question we as a community need to agree
  on (whats our problem statement?) and hope we come out today with
  some progress. - @VonWelch
\item What is the goal of sustainability? Reproducible science,
  persistence, quality, or else?
\item How do you measure success? How do you know when you have
  actually achieved sustainable software?
\item Sustainability should be addressed throughout the life cycle
  (Daisie Huang)
\end{itemize}

\item What is the relationship between sustainability and provenance?

\item Is usability separate from sustainability or a fundamental part?
\end{itemize}

\item from discussion in policy panel

\begin{itemize}
\item Chris Lenhardt's presentation of Venters
  paper~\cite{Venters_WSSSPE} included:

\begin{itemize}
\item What does software sustainability mean
It ``is a rather ambiguous concept...''
\item Lack of an accepted definition gets in the way of integrating
  the concept into software engineering
\item Sustainability as a non-functional requirement
\item Quality of software architectures determines sustainability
\item Propose sustainability is a measure of central quality
  attributes: Extensibility, Interoperability, Maintainability,
  Portability, Reusability, Scalability
\item Develop an architecture evaluation framework based on scenarios
  which help to illuminate how to measure quality/sustainability at
  the architectural level
\end{itemize}

\item Chris Lenhardt's presentation of Katz paper~\cite{Katz_WSSSPE}
  included:
\begin{itemize}
\item Clarifying the notion of sustainability
\begin{itemize}
\item Provide same functionality going forward even if environment changes
\item Is functionality and usability clear
\item Will functionality be correct going into the future even if
  environment changes
\item Is the functionality desired by current and future users
\item Can it incorporate new things as they develop
\end{itemize}
\end{itemize}

\item see slide 9 of
  \url{http://dx.doi.org/10.6084/m9.figshare.853817} for summary table

\item also see discussion at \url{https://docs.google.com/document/d/1eVfioGNlihXG_1Y8BgdCI6tXZKrybZgz5XuQHjT1oKU/edit?pli=1#bookmark=id.2pxw5yey8jl0}

\section{Illustrative Use Cases} \label{sec:use-cases}

@@

\note{Dan: do we need this section?  I think there's some overlap
  between this and the discussion in the Developing and Support
  Software section.  Or perhaps this should go in the cross-cutting
  issues section?}

\section{Conclusions} \label{sec:conclusions}

@@

\todo{say something about licensing - lessons, advice, etc.?}

@@

\begin{tabular}{ L{4cm}  L{4cm}  L{4cm}  L{4cm} }
   Daniel S. Katz
&  Neil Chue Hong
&  Gabrielle Allen
&  David Proctor
\\ Manish Parashar
&  Jay Alameda
&  David Andrs
&  Lorena A Barba
\\ David Bernholdt
&  Phil Bourne
&  Karl Broman
&  Maxine Brown
\\ David Bruhwiler
&  Bruno Bzeznik
&  Jeffrey Carver
&  John W. Cobb
\\ Karen Cranston
&  Rion Dooley
&  Anshu Dubey
&  Ian Foster
\\Juliana Freire
& Derek Gaston
& Derek Groen
& Edmund Hart
\\James Hetherington
& Simon Hettrick
& Kenneth Hoste
& James Howison
\\Daisie Huang
& Kerk Kee
& Arnaud Legrand
& Paul Madden
\\Louis Maddox
& Ketan Maheshwari
& Suresh Marru
& Cezary Mazurek
\\James McClure
& Chris Mentzel
& J. Yates Monteith
& Pierre Neyron
\\Lucas Nussbaum
& Cody Permann
& John Peterson
& Quan Pham
\\Marlon Pierce
& Heather Piwowar
& Jason Riedy
& Andreas Schreiber
\\William Scullin
& Andrew Slaughter
& Arfon Smith
& James Spencer
\\Shel Swenson
& Fr\'{e}d\'{e}ric Suter
& Andy Terrel
& George  Thiruvathukal
\\John Towns
& Erik Trainer
& Matthew Turk
& Eric Van Wyk
\\Colin C. Venters
& Von Welch
& Nancy Wilkins-Diehr
& Theresa Windus
\\Rich Wolski
& Lynn Zentner
& Frank Löffler
& Patrick O'Leary
\end{tabular}

>>>>

\usepackage{color}
\usepackage{dcolumn}
\usepackage{float}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{subfigure}
\usepackage{psfrag}
\usepackage{tabularx}
\usepackage[hyphens]{url}
\usepackage{wrapfig}

@@

@@

\note{papers and presentations should be references - URLs perhaps can just be footnotes for now, but may want to think about this again.}

@@

The First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)\footnote{\url{http://wssspe.researchcomputing.org.uk/WSSSPE1}} was held on Sunday, 17 November 2013, in conjunction with the 2013 International Conference for High Performance Computing, Networking, Storage and Analysis (SC13)\footnote{\url{http://sc13.supercomputing.org}}.

Because progress in scientific research is dependent on the quality and accessibility
of software at all levels, it is now critical to address many challenges related to the
development, deployment, and maintenance of reusable software. In addition, it is
essential that scientists, researchers, and students are able to learn and adopt
software-related skills and methodologies. Established researchers are already
acquiring some of these skills, and in particular a specialized class of software
developers is emerging in academic environments who are an integral and embedded
part of successful research teams. The WSSSPE workshop was intended to provide a
forum for discussion of the challenges, including both positions and experiences.
The short papers and discussion have been archived to provide a basis for continued
discussion, and the workshop has fed into the collaborative writing of this journal publication.
Additional papers based on extended versions of workshop submissions are expected.
The level of interest in the workshop has led the organizers, working with some of the submitters
and some of the attendees, to plan two additional workshops (\todo{details}) and to turn
the workshop website into a community website that can be used as a focus for further
discussion and progress.

\todo{Mention a two-session minisymposium on ``Reliable Computational Science'' in the SIAM Annual Meeting in July. It aims to further explore some of the key issues raised in WSSSPE. Speakers include Tim Davis, Barry Smith, Lorena Barba, Tanu Malik, Bill Rider.}

\todo{mention arXiv:1311.3523 and explain what's different here}

\section{Workshop Process}

WSSSPE1 was organized in collaboration of the relatively small group of \todo{how many} organizers and
a larger peer-review committee consisting of \todo{how many members}. This committee had early influence on the organization,
e.g., on the specific call for papers.

@@

\begin{itemize}
\item the development process that leads to new (versions of) software
\begin{itemize}
\item how fundamental research in computer science or science/engineering domains is turned  into reusable software
\item software created as a by-product of research
\item impact of computer science research on the development of scientific software
\end{itemize}
\item the support and maintenance of existing software, including software engineering
\begin{itemize}
\item governance, business, and sustainability models
\item the role of community software repositories, their operation and sustainability
\end{itemize}
\item the role of open source communities or industry
\item use of the software
\begin{itemize}
\item growing communities
\item reproducibility, transparency needs that may be unique to science
\end{itemize}
\item policy issues, such as
\begin{itemize}
\item measuring usage and impact
\item software credit, attribution, incentive, and reward
\item career paths for developers and institutional roles
\item issues related to multiple organizations and multiple countries, such as intellectual property, licensing, etc.
\item mechanisms and venues for publishing software, and the role of publishers
\end{itemize}
\item education and training
\end{itemize}

@@

58 submissions were received, and almost all submitters used either 
arXiv\footnote{\url{http://arxiv.org}} or 
figshare\footnote{\url{http://figshare.com}} to self-publish their papers.

A peer review process followed the submissions, where the 58 papers received 181 reviews, an 
average of 3.12 reviews per paper. Reviews were completed using a Google form, which allowed 
reviewers to provide scores on relevance and comments to the organizers and to the authors. The review reports enabled the organizers to decide which papers to associate with the workshop, and the authors to improve their papers.

The organizers decided to list 54 of the papers as significantly contributing to the workshop, a very 
high acceptance rate, but one that is reasonable, given the goal of broad participation and the fact 
that the reports were already self-published. The papers were also grouped into three main categories, namely \emph{Developing and Supporting Software}, \emph{Policy}, and \emph{Communities}. Each subject area was associated with a panel and discussion at the workshop.

\section{Agenda}

The workshop consisted of two keynote presentations and three panels/discussion sessions. The panels 
were organized based on a classification of the workshop submissions into three categories, following 
the themes of the call for papers as modified by the areas of the submissions. Four \todo{Is it five?} representatives from 
each submission categories were appointed as panelists, to discuss the papers in that category.

\section{Keynotes}

@@

He noted that in places, development in science has improved thanks to
open source and Github, but for the most part remains arcane. He argues
that there is a lot we can learn from the app-store model. He
mentioned BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
OpenScienceDataCloud \footnote{\url{https://www.opensciencedatacloud.org/}} as
distribution channels.

@@

He experienced problems with staff retention with respect to private organizations which reward 
those combining research and software expertise (the ``Google Bus''). However,
he is a strong fan of private sector sustainability sources and PPP. He
noted that making a successful business from scientific software alone
is very rare: founders overvalue while customers undervalue. He noted
that to last, an open source project needs a minimal funding requirement
even with a vibrant community -- good will only goes so far if you are
being paid to do something else.

He talked about grant schemes of relevance in the U.S., particularly with regard to technology transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}} \todo{change at least first to cite}

@@

\subsection{Scientific Software and the Open Collaborative Web}

@@

His talk used an example from data reduction in Astronomy, where he
needed to remove interfering effects from the device. He built a
``bad pixel mask'', and realized though while this is persistent, there
was no way/practice of sharing this with the data, so many researchers
repeated the same calculations. He estimated that 13 wasted person-years were wasted in
this redundant calculation.

@@

\section{Developing and Supporting Software}

@@

What is sustainable? What are the best practices, can workshops fund meetings to help define and improve best practices in these areas? Software plans from the funding agencies? \todo{mention as cross-cutting, since this came up in other parts of the workshop as well?}

@@

\section{Policy}

@@

\section{Communities}

@@

There were a number of papers categorized under the Communities banner and so two presenters each summarized half of the submissions in this area.

\subsection{Communities Part 1}

This collection of papers was summarized by Karen Cranston.

\subsection{Communities Part 2}

Nancy Wilkins-Diehr summarized 6 papers in "Communities Part 2". The Christopherson paper outlines the degree to which research
relies on high quality software. There are often barriers and a lack of suitable incentives for researchers to embrace software engineering principals.
The Water Science Software Institute is working to lower some of these barriers through an Open Community Engagement Process.
This is a 4-step iterative development process that incorporates Agile development principals.

\item Step 1 - Design - thorough discussion of research questions
\item Step 2 - Develop working code
\item Step 3 - Refine based on new requirements
\item Step 4 - Publish open source

The Christopherson paper reports on the application of Steps 1-3 to a computational modeling framework developed in the 1990s.
Step 1 was realized as a 2-day, in person specifications meeting and code walk-through. Step 2 was a 5-day hack-athon to develop working code.
Step 3 was a 3-day hack-a-thon to refine the code based on new requirements. The team worked on small, low-risk unit of 
code. It was challenging, revealed unanticipated obstacles, programmers had to work together, and experimentation was encouraged.

The paper summarized recommendations to others wishing to engage in this or a similar process. Start small and gradually 
build toward more complex objectives. This is consistent with Agile development. Refactor before adding new functionality.
Approach development as a learning experience. Welcome experimentation, and treat mistakes as a natural part of the learning process.
Repeat Step 1 activities before all hackathons to develop consensus before coding. This allows coding to be the focus of subsequent hackathons. 
In higher risk situations, provide additional time for Step 1 activities. The Christopherson team recommends a minimum of two months.
Ensure any newcomers receive some form of orientation prior to the hackathon, such as a code walkthrough or system documentation.
Co‐locate rather than collaborating remotely whenever feasible.

The Pierce paper described how science gateways can provide a user-friendly entry to complex cyberinfrastructure and they are much
more than a fertile ground for grid computing research. The Pierce paper summarizes the explosion of use of just a few gateways 
over the last 4 years -

@@

\section{Other Discussion}

@@

\section{Cross-cutting Issues}

\subsection{Defining sustainability}

\subsection{Licensing}

\note{not sure if this was really cross-cutting...}

\section{Illustrative use-cases}

@@

\section{Conclusions}

@@

@@

Daniel S. Katz (@danielskatz)

Neil Chue Hong (@npch)

Gabrielle Allen

David Proctor (@djproctor)

Manish Parashar

Jay Alameda (alameda@illinois.edu)

David Andrs

Lorena A Barba (@LorenaABarba)

David Bernholdt (bernholdtde@ornl.gov)

Phil Bourne

Karl Broman (@kwbroman)

Maxine Brown, UIC (maxine@uic.edu)

David Bruhwiler   (bruhwiler@radiasoft.net    @radiasoft)

Bruno Bzeznik

Jeffrey Carver, University of Alabama (carver@cs.ua.edu)

John W. Cobb (cobbjw@ornl.gov, @johnwcobb)

Karen Cranston (@kcranstn, National Evolutionary Synthesis Center)

Rion Dooley

Anshu Dubey (adubey@lbl.gov)

Ian Foster, @ianfoster

Juliana Freire

Derek Gaston

Derek Groen (@whydoitweet on Twitter)

Edmund Hart (@distribecology)

James Hetherington @jamespjh

Simon Hettrick (s.hettrick@software.ac.uk)

Kenneth Hoste (@kehoste, Ghent University, Belgium)

James Howison, University of Texas at Austin (@jameshowison)

Daisie Huang

Kerk Kee (kerk.kee@gmail.com)

Arnaud Legrand

Paul Madden, NOAA / University of Colorado, maddenp@colorado.edu

Louis Maddox (@lmmddx)

Ketan Maheshwari (@km\_9 on Twitter)

Suresh Marru (smarru@iu.edu, @sureshmarru)

Cezary Mazurek, mazurek@man.poznan.pl, @cezary\_mazurek

James McClure

Chris Mentzel @chrismentzel - firstname dot lastname at moore dot org

J. Yates Monteith, Clemson University

Pierre Neyron

Lucas Nussbaum (lucas.nussbaum@loria.fr)

Cody Permann

John Peterson

Quan Pham, University of Chicago

Marlon Pierce, marpierc@iu.edu, @marpierc

Heather Piwowar

Jason Riedy, Georgia Tech, http://fmrl.me/jasonriedy and many other places

Andreas Schreiber (@onyame)

William Scullin

Andrew Slaughter

Arfon Smith

James Spencer (@exp\_n)

Shel Swenson, University of Southern California (shel.swenson@usc.edu)

Fr\'{e}d\'{e}ric Suter

Andy Terrel (@aterrel, andy.terrel@gmail.com, NumFOCUS and Continuum Analytics)

George K. Thiruvathukal, Loyola University Chicago and Computing in Science and Engineering (IEEE CS/AIP) (gkt@cs.luc.edu)

John Towns

Erik Trainer

Matthew Turk (@powersoffour on Twitter, matthewturk@gmail.com)

Eric Van Wyk, University of Minnesota, evw@cs.umn.edu, @ericvanwyk

Colin C. Venters, c.venters@hud.ac.uk

Von Welch, @VonWelch

Nancy Wilkins-Diehr, wilkinsn@sdsc.edu

Theresa Windus (twindus@iastate.edu)

Rich Wolski

Lynn Zentner

Frank Löffler
